Subsidiary Recursion in Coq

### 2022-01-13 paper/paper.tex
added rle example
I think that is all the example recursions I will add.  We have then
just the derivation, interface for subsidiary induction, and examples
of subsidiary induction left to write, and filling in the related
work.  Space is getting tight, so we might need to shrink some earlier
parts a bit.

\section{Introduction: subsidiary recursion}

@@

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy|
(from \verb|Data.List.Extra|) breaks a list into its maximal
sublists whose elements do not satisfy a predicate \verb|p|.  For
example, \verb|wordsBy isSpace " good day "| returns
\verb|["good","day"]|; so \verb|wordsBy isSpace| has the same behavior
as \verb|words| (from \verb|Data.List|).  Code is in
Figure~\ref{fig:wordsBy}. The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination,
but will be permitted by subsidiary recursion as derived below.

\begin{figure}
\begin{verbatim}
wordsBy :: (a -> Bool) -> [a] -> [[a]]
wordsBy p [] = []
wordsBy p (hd:tl) =
  if p hd
  then wordsBy p tl 
  else let (w,z) = span (not . p) tl in
        (hd:w) : wordsBy p z
\end{verbatim}
\caption{Haskell code for \texttt{wordsBy}, demonstrating subsidiary recursion}
\label{fig:wordsBy}
\end{figure}

@@

All presented derivations have been checked with Coq version 8.13.2,
using command-line option \verb|-impredicative-set|.  The code may be
found as release \verb|itp-2022| (dated prior to the ITP 2022
deadline) at \url{https://github.com/astump/coq-subsidiary}.

@@

Figure~\ref{fig:span} gives the algebra \verb|SpanAlg| for computing
\verb|span|, and the code for \verb|spanr|.  We elide the proof
\verb|SpanFunctor| that \verb|SpanF| is indeed a \verb|Functor|,
and make \verb|X| implicit in the constructors of \verb|SpanF|.
The type of \verb|SpanAlg| is
\begin{verbatim}
Alg (ListF A) C SpanF
\end{verbatim}
This states that we are defining an algebra (\verb|Alg|)
for the \verb|ListF A| functor, with anchor type \verb|C| and carrier
\verb|SpanF|.  The algebra does not depend on what the anchor type is,
which is good, as it means we can use \verb|SpanAlg| for top-level or
subsidiary recursions.

Let us continue through the code for \verb|SpanAlg|
(Figure~\ref{fig:span}).  We use \verb|rollAlg| to create an algebra
from something whose type is an application of \verb|AlgF|.  This
takes in all the components of the recursion universe: the abstract
type \verb|R|, the \verb|reveal| function (not needed in this case),
the \verb|fold| function for any subsidiary recursions (also not
needed here), a function we choose to name \verb|span| for making
recursive calls, and finally \verb|xs : ListF A R|.  The algebra
pattern-matches on this \verb|xs|.  In the cases where it is empty or
where its head (\verb|hd|) does not satisfy \verb|p|, we return
\verb|SpanNoMatch|.  This signals to the caller that we really wished
to return \verb|([],xs)|, but could not because we do not have
\verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
we recurse on the tail (\verb|tl|).  This is allowed because
\verb|tl : R| and \verb|span : R -> SpanF R|.  If \verb|span tl|
returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
the suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily,
we have \verb|tl : R| here, so we can do this.  In either case (for
return value of \verb|span tl|), we add the head to the front of the
prefix.  We define \verb|spanhr| to invoke the fold function it is
given, on the algebra (\verb|SpanAlg|).

@@

>>>>

\section{Introduction: subsidiary recursion}
\label{sec:intro}

@@

This paper is about a more expressive form of terminating recursion,
called \textbf{subsidiary recursion}.  While performing an outer
recursion on some input \verb|x|, one may initiate an inner recursion
on \verb|x| (or possibly some of its subdata), preserving the
possibility of further invocations of the outer recursive function.
Let us see a simple example.  The function \verb|wordsBy| (from
\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
elements do not satisfy a predicate \verb|p|.  For example,
\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|; so
\verb|wordsBy isSpace| has the same behavior as \verb|words| (from
\verb|Data.List|).  Code is in Figure~\ref{fig:wordsBy}, using
also \verb|break| (\verb|Data.List|). Recall that
\verb|break p| is equivalent to \verb|span (not . p)|: it splits its
input into the maximal prefix whose elements do not satisfy \verb|p|,
and the following suffix.  The first recursive call,
\verb|wordsBy p tl|, is structural.  But in the second, we invoke
\verb|wordsBy p| on a value obtained from another recursion, namely
\verb|span|.  This is not allowed under structural termination, but
will be permitted by subsidiary recursion as derived below.

\begin{figure}
\begin{verbatim}
wordsBy :: (a -> Bool) -> [a] -> [[a]]
wordsBy p [] = []
wordsBy p (hd:tl) =
  if p hd
  then wordsBy p tl 
  else let (w,z) = break p tl in
        (hd:w) : wordsBy p z
\end{verbatim}
\caption{Haskell code for \texttt{wordsBy}, demonstrating subsidiary recursion}
\label{fig:wordsBy}
\end{figure}

@@

All presented derivations have been checked with Coq version 8.13.2,
using command-line option \verb|-impredicative-set|.  The code may be
found as release \verb|itp-2022| (dated prior to the ITP 2022
deadline) at \url{https://github.com/astump/coq-subsidiary}.  The
paper references files in this codebase, as an aid to the reader
wishing to peruse the code.

@@

Figure~\ref{fig:span} gives the algebra \verb|SpanAlg| for computing
\verb|span|, and the code for \verb|spanr|.  We elide the proof
\verb|SpanFunctor| that \verb|SpanF| is indeed a \verb|Functor|,
and make \verb|X| implicit in the constructors of \verb|SpanF|.
The type of \verb|SpanAlg p C| is
\begin{verbatim}
Alg (ListF A) C SpanF
\end{verbatim}
This states that we are defining an algebra (\verb|Alg|) for the
\verb|ListF A| functor, with anchor type \verb|C| and carrier
\verb|SpanF|.  \verb|SpanF| has type \verb|Set -> Set|, as required
for the carriers of our algebras. The definition of \verb|SpanAlg| is
actually parametrized by \verb|C|, which is good, as it means we can
use \verb|SpanAlg| for top-level or subsidiary recursions.

Let us continue through the code for \verb|SpanAlg|
(Figure~\ref{fig:span}).  We use \verb|rollAlg| to create an algebra
from something whose type is an application of \verb|AlgF|.  This
takes in all the components of the recursion universe: the abstract
type \verb|R|, the \verb|reveal| function (not needed in this case),
the fold function (\verb|fo|) for any subsidiary recursions (also not
needed here), a function we choose to name \verb|span| for making
recursive calls, and finally \verb|xs : ListF A R|.  The algebra
pattern-matches on this \verb|xs|.  In the cases where it is empty or
where its head (\verb|hd|) does not satisfy \verb|p|, we return
\verb|SpanNoMatch|.  This signals to the caller that we really wished
to return \verb|([],xs)|, but could not because we do not have
\verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
we recurse on the tail (\verb|tl : R|) by calling the provided \verb|span : R -> SpanF R|.
If \verb|span tl|
returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
the suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily,
we have \verb|tl : R| here, so we can do this.  In either case (for
return value of \verb|span tl|), we add the head to the front of the
prefix.  We define \verb|spanhr| to invoke the fold function it is
given, on the algebra (\verb|SpanAlg|).

@@

We can easily define \verb|break|, in Figure~\ref{fig:break}.  The
function \verb|breakr| is a version of \verb|break| that can be used
for subsidiary recursion, similarly to \verb|spanr| for \verb|span|.
Such a function always takes in a fold function (\verb|fo|) with
anchor type \verb|R|, which then is used to fold the algebra in
question.

\begin{figure}
\begin{verbatim}
Definition breakr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                 (p : A -> bool)(xs : R) : list A * R :=
  spanr fo (fun x => negb (p x)) xs.

Definition break(p : A -> bool)(xs : List A) : list A * List A :=
  breakr (fold (ListF A)) p xs.
\end{verbatim}
\caption{The \texttt{break} function and its more flexible version, \texttt{breakr}, defined in terms of \texttt{spanr} (Figure~\ref{fig:span})}
\label{fig:break}
\end{figure}

\subsection{The \texttt{wordsBy} function (\texttt{WordsBy.v})}

Let us now see how to write \verb|wordsBy|, our example function from
Section~\ref{sec:intro}, using \verb|spanr| as a subsidiary recursion.
The code is in Figure~\ref{fig:wordsby}, assuming a type
\verb|A : Set|.  The setup is similar to that for \verb|span|.  We first define
an algebra \verb|WordsBy|, parametrized by anchor type \verb|C| (and also
the predicate \verb|p|), of type
\begin{verbatim}
Alg (ListF A) C (Const (list (list A)))
\end{verbatim}
\noindent This says that \verb|WordsBy p C| is an algebra (\verb|Alg|)
for the \verb|ListF A| functor, with anchor type \verb|C|, and carrier
\verb|Const (list (list A))|.  \verb|Const| is the combinator for
creating the object part of constant functors; \verb|FunConst| creates
the morphism part (i.e., the \verb|fmap| function).  We use it
\verb|Const| here and in other examples where the return type of the
algebra will not depend on its abstract type \verb|R|.  Here, we are
constructing from scratch a list of lists, so it will not be legal to
recurse on the list itself, or its (list) elements.  Intead, we just
use the \verb|list| type of Coq's standard library.

The code for \verb|WordsBy| is, except for the noise of \verb|rollAlg|
and accepting the components of the recursion universe, essentially
the same as what we saw in Section~\ref{sec:intro}.  We pattern match
on \verb|xs : ListF A R|.  Recall that for this function, we are
trying to drop elements which satisfy \verb|p|, and return a list of
the sublists between maximal sequences of such elements.  In the
\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we are
supposed to drop it and recurse.  This is legal, because \verb|tl : R|
and \verb|wordsBy : R -> list (list A)|.  In the \verb|else| case,
we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
does not satisfy \verb|p|, and the remaining suffix \verb|z|.

Here we see the benefit of our approach.  From Figure~\ref{fig:break},
the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
is the anchor type of the provided fold function \verb|fo|.  And
\verb|fo| has type \verb|FoldT (ListF A) Alg R|, from the definition
of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
with \verb|ListF A|).  This means that from the invocation of
\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
can indeed apply \verb|wordsBy : R -> list (list A)| to
\verb|z| to recurse.

\begin{figure}
\begin{verbatim}
  Definition WordsBy(p : A -> bool)(C : Set)
    : Alg (ListF A) C (Const (list (list A))) :=
    rollAlg (fun R reveal fo wordsBy xs => 
         match xs with
           Nil => [] 
         | Cons hd tl =>
           if p hd then
             wordsBy tl
           else
             let (w,z) := breakr fo p tl in
             (hd :: w) :: wordsBy z
         end).

Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                   (p : A -> bool)(xs : R) : list (list A) :=
  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p R) xs.

Definition wordsBy(p : A -> bool)(xs : List A) : list (list A) :=
  wordsByr (fold (ListF A)) p xs.
\end{verbatim}
\caption{The \texttt{wordsBy} and \texttt{wordsByr} function, defined using an algebra}
\label{fig:wordsby}
\end{figure}

\subsection{The \texttt{mapThrough} function (\texttt{MapThrough.v})}

The Haskell library \verb|Data.List.Extra| has a function \verb|repeatedly|, which
is defined essentially as follows; I have attempted a more informative name:
\begin{verbatim}
mapThrough :: (a -> [a] -> (b, [a])) -> [a] -> [b]
mapThrough f [] = []
mapThrough f (a:as) = b : mapThrough f as'
    where (b, as') = f a as
\end{verbatim}
\noindent The idea is that the function is like the standard \verb|map| function
on lists, except that here, the function \verb|f| that we are mapping (or ``mapping
through'') takes in not just the current element \verb|a|, but also the tail \verb|as|.
It then returns the value \verb|b| to include in the output list, and whatever other
list it wishes, upon which \verb|mapThrough| will recurse.

We can write this combinatory using our infrastructure for subsidiary recursion.
For this to work, we need to supply the mapped function with the fold function
for \verb|mapThrough|'s recursion.  This is so that the mapped function can
initiate a subsidiary recursion, returning a value in the abstract type \verb|R|
of \verb|mapThrough|'s recursion.  So the type we will use for mapped functions
is:
\begin{verbatim}
Definition mappedT(A B : Set) : Set :=
  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
\end{verbatim}
\noindent This type is more informative than the Haskell type,
since it shows that the second component of the returned value
must have type \verb|R|, and hence must be (hereditarily) a tail
of the input of type \verb|R|.

Given this definition, the code for \verb|mapThrough| and
\verb|mapThroughr| is in Figure~\ref{fig:mapthrough}.  The code for
\verb|MapThroughAlg| is very similar (discounting syntax) to the
Haskell code above.  Here, though, when we call \verb|f|, we must
supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
from the definition of \verb|mappedT|, we have that \verb|b : B| and
\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> List B|
on \verb|c|.  Note that as we are building up a new list from scratch
(rather than just extracting some tail of the input list), we just
return \verb|list B|; we cannot perform further subsidiary recursion
on the output.

\begin{figure}
\begin{verbatim}
Definition MapThroughAlg{B : Set}(f:mappedT A B)
           (C : Set) : Alg (ListF A) C (Const (list B)) :=
  rollAlg (fun R reveal fo mapThrough xs => 
    match xs with
      Nil => []
    | Cons hd tl =>
      let (b,c) := f R fo hd tl in
        b :: mapThrough c
    end).

Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                      {B : Set}(f:mappedT A B) : R -> list B :=
  fo (Const (list B)) (FunConst (list B)) (MapThroughAlg f R).

Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B :=
  mapThroughr (fold (ListF A)) f.
\end{verbatim}
\caption{The \texttt{mapThrough} and \texttt{mapThroughr} functions,
  with their defining algebra}
\label{fig:mapthrough}
\end{figure}

\subsection{Run-length encoding}

Using \verb|mapThrough|, we can write a quite concise function for
\emph{run-length encoding}, a basic data-compression algorithm where
maximal sequences of $n$ occurrences of element $e$ are summarized by
the pair $(n,e)$~\cite{datacomp}.  In Haskell, invoking \verb|span| and
\verb|mapThrough| (defined above), the code is simply
\begin{verbatim}
rle :: Eq a => [a] -> [(Int,a)]
rle = mapThrough compressSpan
  where compressSpan a as =
          let (p,s) = span (== a) as in
            ((1 + length p, a),s)
\end{verbatim} 
\noindent (Recall that \verb|(== a)| is a Haskell \emph{section}
testing its input for equality with \verb|a|.)  The
\verb|compressSpan| helper function gathers up all elements at the
start of the tail \verb|as| that are equal to the head \verb|a|.  This
prefix is returned as \verb|p|, with the remaining suffix as \verb|s|.
The pair \verb|(1 + length p, a)| is returned to summarize
\verb|a :: p|.  We then use \verb|mapThrough| to iterate
\verb|compressSpan| through the suffix \verb|s|.

Assuming \verb|A : Set| and an equality test \verb|eqb : A -> A -> bool| on it,
code for run-length encoding using our infrastructure is listed in Figure~\ref{fig:rle}.
The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
\verb|compressSpan| has type:
\begin{verbatim}
forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> (nat * A) * R.
\end{verbatim}
\noindent It will be invoked by the code for \verb|mapThrough| with a
fold function \verb|fo| with anchor type \verb|R|, and then has the
responsibility of mapping the tail at type \verb|R| (second input) to
a result upon which \verb|mapThrough| should recurse (second component
of the output pair).  Then we define an algebra \verb|RleAlg| by
supplying \verb|compressSpan| as the function to map through, to
\verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).  Following the
pattern we have seen in all the examples above, we may then define
function \verb|mapThroughr| for subsidiary recursions, and
\verb|mapThrough| for top-level recursions.

\begin{figure}
\begin{verbatim}
Definition compressSpan : mappedT A (nat * A) :=
  fun R fo hd tl => 
    let (p,s) := spanr fo (eqb hd) tl in
       ((succ (length p),hd), s).

Definition RleCarr := Const (list (nat * A)).
Definition RleAlg(C : Set) : Alg (ListF A) C RleCarr :=
  MapThroughAlg compressSpan C.

Definition rle(xs : List A) : list (nat * A)
  := @fold (ListF A) RleCarr (FunConst (list (nat * A))) (RleAlg (List A)) xs.
\end{verbatim}
\caption{The function \texttt{rle} for run-length encoding, and the algebra \texttt{RleAlg} defining it
in terms of \texttt{MapThroughAlg} (Figure~\ref{fig:mapthrough})}
\label{fig:rle}
\end{figure}
